---
title: "wram_working_memory_analysis"
author: "Nick Fox"
date: "Tuesday, March 03, 2015"
output: html_document
---

This is the first pass at analyzing the wRAM working memory protocol in R.

```{r echo = FALSE}

library(dplyr)
library(ggplot2)
library(broom)



setwd("C:/Users/Nick/Documents/GitHub/wram_working_memory")

wram_raw <- read.csv("wram_wm_raw.csv", header = TRUE, sep = ",")

wram_raw <- tbl_df(wram_raw)

wram_cleaned <- na.omit(wram_raw)

wram_averages <- wram_cleaned %>%
  dplyr::group_by(Day, Treatment, Trial) %>%
  dplyr::mutate(avr_RM_error = mean(RM_error), avr_WMC_error = mean(WMC_error)) %>%
  distinct(avr_RM_error, avr_WMC_error)
#generates average number of RM and WMC errors per trial per treatment
```

```{r echo = FALSE}
wram_averages %>%
  filter(Trial != 1) %>%
  ggplot(aes(x = Trial, y = avr_WMC_error, group = Treatment, colour = Treatment)) + geom_point(size = 3) + geom_line(size = 1, alpha = 0.8) + facet_grid(. ~ Day) + theme_bw() + scale_x_continuous(breaks = seq(2, 4, 1), limits = c(1.5, 4.5)) + ylab("Working Memory Correct Errors") + scale_color_brewer(palette="Dark2")
#graph of working memory correct errors over trials, faceted by day.
```

```{r echo = FALSE}
wram_averages %>%
  filter(Trial != 1) %>%
  ggplot(aes(x = Trial, y = avr_RM_error, group = Treatment, colour = Treatment)) + geom_point(size = 3) + geom_line(size = 1, alpha = 0.8) + facet_grid(. ~ Day) + theme_bw() + scale_x_continuous(breaks = seq(2, 4, 1), limits = c(1.5, 4.5)) + ylab("Reference Memory Errors") + scale_color_brewer(palette="Dark2")
#graph of reference memory errors over trials, faceted by day.
```


```{r echo = FALSE}

wram_averages_treatment <- wram_averages %>%
  filter(Day != 1) %>%
  dplyr::group_by(Day, Treatment) %>%
  dplyr::mutate(treatment_RM = mean(avr_RM_error), treatment_WMC = mean(avr_WMC_error)) %>%
  distinct(treatment_RM, treatment_WMC)
#creates the data frame to make the following graph (for average number of errors per treatment per day).

wram_averages_treatment %>%
  ggplot(aes(x = Day, y = treatment_WMC, group = Treatment, colour = Treatment)) + geom_point(size = 3) + geom_line(size = 1, alpha = 0.8) + geom_smooth(method = "lm", se = FALSE) + theme_bw() + scale_x_continuous(breaks = seq(2, 12, 1)) + ylab("Working Memory Correct Errors") + scale_color_brewer(palette="Dark2")

aov.out.WMC <- aov(treatment_WMC ~ Treatment * Day + Error(Animal), data = wram_averages_treatment)
summary(aov.out.WMC)

```


```{r echo = FALSE}

wram_averages_treatment %>%
  ggplot(aes(x = Day, y = treatment_RM, group = Treatment, colour = Treatment)) + geom_point(size = 3) + geom_line(size = 1, alpha = 0.8) + geom_smooth(method = "lm", se = FALSE) + theme_bw() + scale_x_continuous(breaks = seq(2, 12, 1)) + ylab("Reference Memory Errors") + scale_color_brewer(palette="Dark2")
#graph of reference memory errors over day, averaged by treatment.

aov.out.RM <- aov(treatment_RM ~ Treatment * Day + Error(Animal), data = wram_averages_treatment)
summary(aov.out.RM)
```


```{r echo = FALSE}
setwd("C:/Users/Nick/Documents/GitHub/single_platform_wram")
wram_raw <- read.csv("one_platform_wram_raw.csv", header = TRUE)

se <- function(x) sqrt(var(x)/length(x))
#adds a function to make standard error calculations (for error bars)

wram_raw_male <- wram_raw %>%
  filter(Sex != "Female")
#makes male only dataframe

wram_raw_female <- wram_raw %>%
  filter(Sex != "Male")
#makes female only dataframe

wram_raw_female2 <- wram_raw_female %>%
  dplyr::group_by(Day, Animal) %>%
  dplyr::mutate(avr.latency = mean(Duration), avr.distance = mean(Distance)) %>%
  distinct(avr.latency, avr.distance)
#create a new data frame with average latency and average distance travelled per female per day (averaged trials)

wram_raw_male2 <- wram_raw_male %>%
  dplyr::group_by(Day, Animal) %>%
  dplyr::mutate(avr.latency = mean(Duration), avr.distance = mean(Distance)) %>%
  distinct(avr.latency, avr.distance)
#create a new data frame with average latency and average distance travelled per male per day (averaged trials)

wram_raw_female3 <- wram_raw_female2 %>%
  dplyr::group_by(Day, Treatment) %>%
  dplyr::mutate(mean.treatment.latency = mean(avr.latency), sd.treatment.latency = sd(avr.latency), sem.treatment.latency = se(avr.latency), mean.treatment.distance = mean(avr.distance), sd.treatment.distance = sd(avr.distance), sem.treatment.distance = se(avr.distance))
#made a new dataframe with mean, std, and SEM calculated by animal x day and by treatment x day
##avr.latency and avr.distance are animal by day. ##mean.latency and mean.distance are treatment

wram_raw_male3 <- wram_raw_male2 %>%
  dplyr::group_by(Day, Treatment) %>%
  dplyr::mutate(mean.treatment.latency = mean(avr.latency), sd.treatment.latency = sd(avr.latency), sem.treatment.latency = se(avr.latency), mean.treatment.distance = mean(avr.distance), sd.treatment.distance = sd(avr.distance), sem.treatment.distance = se(avr.distance))
#made a new dataframe with mean, std, and SEM calculated by animal x day and by treatment x day
##avr.latency and avr.distance are animal by day. ##mean.latency and mean.distance are treatment

```


This next set of analyses comes from averaging the data by treatment per day.  Each treatment (SEA or Saline) has 1 point per day.  That point is the average of all of the measures made for the animals of that treatment that day.

#Females - Average Distance Traveled

```{r echo = FALSE}

wram_raw_female3 %>%
  group_by(Day, Treatment) %>%
  ggplot(aes(x = Day, y = mean.treatment.distance, group = Treatment, colour = Treatment)) + ylab("Average distance travelled (m)")+ geom_point(size = 4) + geom_line(size = 1) + scale_color_brewer(palette="Dark2") + theme_bw()
#graph the average distance by treatment

aov.out.F.avrdistance <- aov(avr.distance ~ Treatment * Day + Error(Animal), data = wram_raw_female2)
summary(aov.out.F.avrdistance)
#repeated measures anova for distance (females) using average distance by treatment

```
#Females - Average latency to find platform

```{r echo = FALSE}

wram_raw_female3 %>%
  group_by(Day, Treatment) %>%
  ggplot(aes(x = Day, y = mean.treatment.latency, group = Treatment, colour = Treatment)) + ylab("Average latency to find platform (s)") + geom_point(size = 4) + geom_line(size = 1) + scale_color_brewer(palette="Dark2") + theme_bw()
#graph the average latency by treatment

aov.out.F.avrlatency <- aov(avr.latency ~ Treatment * Day + Error(Animal), data = wram_raw_female2)
summary(aov.out.F.avrlatency)
#repeated measures anova for duration (females) using average latency by treatment
```

#Males - Average Distance Traveled

```{r echo = FALSE}

wram_raw_male3 %>%
    group_by(Day, Treatment) %>%
    ggplot(aes(x = Day, y = mean.treatment.distance, group = Treatment, colour = Treatment)) + ylab("Average distance travelled (m)")+ geom_point(size = 4) + geom_line(size = 1) + scale_color_brewer(palette="Dark2") + theme_bw()
#graph the average distance by treatment

aov.out.M.avrdistance <- aov(avr.distance ~ Treatment * Day + Error(Animal), data = wram_raw_male2)
summary(aov.out.M.avrdistance)
#repeated measures anova for distance (females) using average distance by treatment
```

#Males - Average latency to find platform

```{r echo = FALSE}

wram_raw_male3 %>%
    group_by(Day, Treatment) %>%
  ggplot(aes(x = Day, y = mean.treatment.latency, group = Treatment, colour = Treatment)) + ylab("Average latency to find platform (s)") + geom_point(size = 4) + geom_line(size = 1) + scale_color_brewer(palette="Dark2") + theme_bw()
#graph the average latency by treatment

aov.out.M.avrlatency <- aov(avr.latency ~ Treatment * Day + Error(Animal), data = wram_raw_male2)
summary(aov.out.M.avrlatency)
```





```{r}
wram_raw_female3 <- wram_raw_female2 %>%
  dplyr::group_by(Day, Treatment) %>%
  dplyr::mutate(mean.latency = mean(avr.latency), sd.latency = sd(avr.latency), sem.latency = se(avr.latency), mean.distance = mean(avr.distance), sem.distance = se(avr.distance))
```
